---
title: "Intro to Data Frames in R"
subtitle: "Environmental Data Analytics"
author: "Luana Lima"
date: "Fall 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length.

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

```{r cars}
cars
```

Note that it has 2 columns and 50 rows.

## Data frame columns

Suppose you want just the column speed. How would you access that data?

```{r}
cars$speed
```

How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```

## Transforming object in a data frame

Is the new object you create a data frame?

```{r}
#Option 1
class(car_speed)

#Option 2
is.data.frame(car_speed)
```

How could you make it a data frame?

```{r}
df_car_speed <- as.data.frame(car_speed)
df_car_speed
class(df_car_speed)
```

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

```{r}
car_dist <- cars$dist

#Option 1
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind()
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist)
class(df_opt2)
df_opt2

```

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names()
names(df)
```

## Creating a data frame

How would you create a data frame?

```{r}
#useful function data.frame()
create_df <- data.frame("speed"=car_speed,"dist"=car_dist)
create_df
```

## Data frame functions

Some useful functions to use with data frames.

```{r naming the chunk}
ncol(df)  
nrow(df)  
length(df)  #same as ncol
summary(df)
head(df) #show the first 6 rows of df

#If you know the number of the column you want you can refer to that to access column
df[,1]

#you could also use this notation to delete columns
df <- df[,-2]
df
```



Notes
- Tidy data: Observable (horizontal), variables(vertical), values

-csv file is universal can be read everywhere, excel is binary

Data structures
- vector (one dimensional) and matrix (more dimensions make it matrix, 2 columns) are arrays
- list; mix and match data types, number string or list but with a vector if you put numbers --> numeric
Vector can only be one type.

include = FALSE or TRUE
echo = TRUE or FALSE

R Coding:
- Values: 1,2,cat, Ivan, drum, FALSE
-Objects: naming thing, c ("Red","Blue") colors vector
-Functions: mean (2,5)

data.frame (from scratch) vs as.data.frame (convert something to a data frame)

cbind doesnt let you change names of the column, data frame does.

A: unstaged file, no changes have been made
M: edited file 
Red: commit the delete, git has the file but R doesnt 
Yellow: files haven't been committed at all

