---
title: "Assignment 2: Coding Basics"
author: "Sakina Shahid"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq_30 <- seq(1,30,3)
#the code above asks R to take the sequence from 1 to 30 by a space of 3

#2. 

mean (seq_30)
median (seq_30)

# here I am computing the mean and median of the entire sequence using inbuilt R functions

mean <- mean(seq_30)
median <- median(seq_30)

#I am creating variable names

#3. 
mean > median

#the statement asks whether the mean is greater than the median. The output is FALSE indicating that it is not greater.

```


## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

names_of_students <- c("Sakina","Umi","Ivan","Abbas")
#names of students is a character vector 
test_scores <- c (10,25,75,90)
#test scores are a numerical vector
pass_fail <- c(FALSE,FALSE,TRUE,TRUE)
#pass fail assigns true or false to numerical values, it is a logical class
class(pass_fail)


```
```{r}
#Step 1: Converting my vectors to data frames
names_of_students.df <- as.data.frame(names_of_students)
is.data.frame(names_of_students.df)

test_scores.df <- as.data.frame(test_scores)
is.data.frame(test_scores.df)

pass_fail.df <- as.data.frame(pass_fail)
class(pass_fail.df)
```
```{r}

school.df <- data.frame("Name"=names_of_students,"Score"=test_scores,"Pass"=pass_fail)

class(school_final.df)
school.df

#I named the data frame school since all of the vectors fall under this category



```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r}

school_scores <- function(x){
  umi <- ifelse (test_scores>50,"PASS","FAIL")
  return(umi)
}
```

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

school_is_cool <- school_scores(test_scores)
print(school_is_cool)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
