---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Sakina Shahid"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

library(tidyverse)
library(agricolae)   
library(here)
library(lubridate)

ntl.df <- read.csv(here('Data','Raw',                  'NTL-LTER_Lake_ChemistryPhysics_Raw.csv'),stringsAsFactors = TRUE)
ntl.df$sampledate <- mdy(ntl.df$sampledate)
class(ntl.df$sampledate)

#2

theme <- theme_linedraw(base_size = 13) +
  theme(axis.text = element_text
    (color = "black"),plot.title = element_text(color="black",size=12),
    axis.title=element_text(size=10),
plot.background=element_rect(fill="lightyellow"), 
panel.border=element_rect(linewidth = 1.5,linetype = "solid"),
axis.ticks= element_line(color="darkgrey",size=2),
panel.grid.major = element_line(size=0.1))

theme_set(theme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no change in mean lake temperature with depth across all lakes in July.
Ha: There is change in mean lake temperature with depth across all lakes in July.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ntl_final.df <- ntl.df %>%
  filter(month(sampledate)==7)%>%
  select(lakename,year4,daynum,depth,temperature_C) %>%
  na.omit()

#5

scatterplot <- ggplot(ntl_final.df,aes(x=depth,
                      y=temperature_C))+ 
                      geom_point(size=0.9) + 
                      geom_smooth(method=lm) +
                      ylim(c(0, 35))+
                      ggtitle("Temperature vs Depth for Lakes in July")+
                      ylab("Temperature (C)")+
                      xlab("Depth")
                  
                    print(scatterplot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure above suggests that as depth increases, temperature decreases. While there is a downward trend, there is a sharp drop in temperature at a depth of 7m and the data points are clustered closer together in relation to the negative linear slope.There is a high deviation of data points from the slope at shallower depths.  


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

ntl.regression <- 
  lm(ntl_final.df$temperature_C ~ 
       ntl_final.df$depth)
summary(ntl.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model with depth has an adjusted r sqaured value of 0.7387 indicating that approximately 73.87% of temperature change can be explained by depth.There is a large number for degrees of freedom indicating that there are 9726 independent values from which the result is generated, resulting is more statistical power.For every 1m increase in depth, there is a -1.9 C decrease in temperature.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

model_1 <- lm(data = ntl_final.df, temperature_C ~ year4 + daynum + 
              depth)
step(model_1)

#10
summary(model_1)

```



11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests using all the variables - daynum, year and depth to predict temperature. This model explains 0.7411 or 74.11% of the change in temperature compared to the r sqaured value of 0.7387 in the depth only model. It is a slight improvement over the model using only depth, indicating that depth is most likely the strongest indicator of changing temperature.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

temperature.anova <- aov(data = ntl_final.df, temperature_C ~ lakename)
summary(temperature.anova)

temperature.lm <- lm(data = ntl_final.df, temperature_C ~ lakename)
summary(temperature.lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes. The model has a p value of less than 0.05 approaching 0. While this shows there is significant difference in mean temperature, it does not tell us between what lakes. The model also has a high degree of freedom, indicating a large sample size lending to more statistical power.Based on this analysis, we can reject the null hypothesis. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

plot1 <- ggplot(ntl_final.df,
    aes(x=depth,y=temperature_C,color=lakename))+ 
  geom_smooth(method=lm,se=FALSE) + 
  geom_point(alpha=0.5) + 
  ylim(c(0,35)) + 
  ggtitle("Lakes and Changes in Temperature with Depth")

print(plot1)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(temperature.anova)

lake.groups <- HSD.test(temperature.anova, "lakename", group = TRUE)
lake.groups


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter, Paul and Ward Lake have the same temperature statistically speaking. Based on the groupings, there is no lake that has a mean temperature that is statistically distinct from all the other lakes.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can do a two sample t test to test whether the mean temperatures are distinct. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
final.df <- ntl_final.df %>% 
            filter(lakename=="Crampton Lake"|lakename=="Ward Lake")

twosample <- t.test(final.df$temperature_C ~ final.df$lakename)
twosample
```

>Answer: Based on the p value of 0.2649 which is greater than 0.05, we cannot reject the null hypothesis. This indicates that there is no statistically significant difference between the temperature in Crampton Lake and Ward Lake for the month of July.This matches my answer for Q16 because Crampton Lake has the initials "ab" and Ward Lake has the initials "bc".
